Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2019-10-17 manuscript/manuscript.Rmd
nates edits

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and state extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e. scaled anomalies) calculated on a per-station basis. In Rhode Island, temperature and chlorophyll a increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time. In short, algla biomass, as measured by cholorphyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Although not causal, this analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
---

@@

Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, scale does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g. northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*.  NEED TO ADD IN SENTENCE OR TWO ABOUT WHAT COULD BE DRIVING INCRESE (WARMING OR OTHER STUFF WE DIDN"T LOOK AT). Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

In addition to the annualized trends we address with this study, there are other trends that may be of interest.  For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.  Anecdotal evidence in Rhode Island points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.  For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?  These questions are beyond the scope of this study, but all warrant further, careful investigation.

@@

There are several possible approaches to further reduce nutrient loads [@yang2018nutrients]. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].  Additionally, within-lake approaches may also be used.  Another approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].

@@

The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g. @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

>>>>

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and state extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e. scaled anomalies) calculated on a per-station basis. In Rhode Island, temperature and chlorophyll a increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time. In short, algla biomass, as measured by cholorphyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Although not causal, this analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
---

@@

Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

@@

Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g. northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.  For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.  Anecdotal evidence in Rhode Island points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.  For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status? Lastly, changes in rainfall, extreme weather events, or other climate mediated factors can also be playing a role in increasing chlorophyll in Rhode Island lakes and reservoirs. These questions are beyond the scope of this study, but all warrant further, careful investigation.

@@

There are several possible approaches to further reduce nutrient loads [@yang2018nutrients]. First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.  Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].  Additionally, within-lake approaches may also be used.  Another approach is the restoration of freshwater mussels to waterbodies that historically had those species. Some studies using freshwater mussels have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].

@@

The analysis approach we used here, site-specific z-scores, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g. @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
